---
title: "Introduction to jmjax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jmjax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What jmjax fits

A shared-parameter joint model links two submodels through a common
random effect:

- **Longitudinal submodel**: a linear mixed-effects model,
  `y_ij = X_ij %*% beta + Z_ij %*% b_i + e_ij`.
- **Survival submodel**: a relative-risk (proportional hazards) model,
  `h_i(t) = h0(t) * exp(alpha * m_i(t))`, where `m_i(t)` is the subject's
  fitted longitudinal trajectory at time `t`, and `alpha` (the
  **association parameter**) measures how strongly the longitudinal
  trajectory predicts the hazard.

jmjax offers two estimation approaches, each with its own function, and
two choices for the baseline hazard `h0(t)`:

| | Weibull baseline | Spline baseline |
|---|---|---|
| **Maximum likelihood** (adaptive Gauss-Hermite) | `jm_mle(baseline = "weibull")` | `jm_mle(baseline = "spline")` |
| **Bayesian** (NUTS) | `jm_bayes(baseline = "weibull")` | `jm_bayes(baseline = "spline")` |

Both functions take the same model arguments. Each documents only the
options that apply to it (optimizer and quadrature settings for
`jm_mle()`; chains, priors and sampler settings for `jm_bayes()`).
`jm_fit()`, which both call, remains available with the method names
`"weibull-PH-aGH"`, `"spline-PH-aGH"`, `"weibull-PH-mcmc"` and
`"spline-PH-mcmc"`.

## One-time setup

```{r, eval = FALSE}
library(jmjax)
jmjax_setup()  # creates a managed Python virtualenv with jax/numpyro
```

If you already have a Python environment with `jax`/`numpyro` installed
(e.g. a conda environment), point `reticulate` at it instead - `jmjax`
will use it automatically and skip creating its own:

```{r, eval = FALSE}
library(reticulate)
use_condaenv("my_existing_env", required = TRUE)
library(jmjax)  # will detect and use the active environment
```

## Example data

The examples use the primary biliary cirrhosis data from the `JM`
package: serum bilirubin measured repeatedly during follow-up (`pbc2`,
one row per visit) and the time to death (`pbc2.id`, one row per
patient). The two data frames share an `id` column, which is all jmjax
needs to link them. `status2` is 1 for death and 0 otherwise.

The output shown in this vignette comes from a real run of this code
(`dev/intro_vignette_outputs.R` in the source repository). The code is
not run when the vignette is built, because that needs the Python
backend. MCMC results vary slightly between machines.

```{r, eval = FALSE}
library(survival)
data("pbc2", "pbc2.id", package = "JM")
pbc2$log_bili <- log(pbc2$serBilir)

head(pbc2[, c("id", "year", "log_bili", "drug")], 4)
#>   id      year    log_bili      drug
#> 1  1 0.0000000  2.67414865 D-penicil
#> 2  1 0.5256817  3.05870707 D-penicil
#> 3  2 0.0000000  0.09531018 D-penicil
#> 4  2 0.4983025 -0.22314355 D-penicil
head(pbc2.id[, c("id", "years", "status2", "drug")], 4)
#>   id     years status2      drug
#> 1  1  1.095170       1 D-penicil
#> 2  2 14.152338       0 D-penicil
#> 3  3  2.770781       1 D-penicil
#> 4  4  5.270507       1 D-penicil
```

## Fitting: the fast first look

Maximum likelihood with a Weibull baseline hazard is the fastest option,
has no priors to configure, and is a good first look at your data or a
source of starting values for a more flexible fit later:

```{r, eval = FALSE}
fit_w <- jm_mle(log_bili ~ year, Surv(years, status2) ~ drug,
                data_long = pbc2, data_surv = pbc2.id,
                id_var = "id", time_var = "year", baseline = "weibull")
fit_w
#> Call:
#> jm_mle(long_formula = log_bili ~ year, surv_formula = Surv(years, 
#>     status2) ~ drug, data_long = pbc2, data_surv = pbc2.id, id_var = "id", 
#>     time_var = "year", baseline = "weibull")
#>
#> Method: weibull-PH-aGH 
#> Precision: float64
#> Subjects: 312  | Longitudinal obs: 1945  | Events: 140 (44.9%) 
#>
#> Association (alpha): effect of the longitudinal marker on the hazard
#>   alpha = 1.2797, 95% CI = [1.0709, 1.4884]
#>   hazard ratio per unit of marker = 3.5954, 95% CI = [2.9181, 4.4299]
#>
#> Estimates by submodel:
#> Longitudinal:
#>         Estimate Std.Err z.value p.value
#> beta_0    0.5775  0.0650  8.8894 <0.0001
#> beta_1    0.0981  0.0043 22.7496 <0.0001
#> sigma_e   0.4913  0.0086 57.2099 <0.0001
#>
#> Survival (baseline hazard):
#>             Estimate Std.Err  z.value p.value
#> log_lambda0  -4.5450  0.2933 -15.4953 <0.0001
#> shape         1.1994  0.0873  13.7416 <0.0001
#>
#> Survival (baseline covariates):
#>         Estimate Std.Err z.value p.value
#> gamma_0   0.0219  0.1754  0.1247   0.901
#>
#> Variance components:
#>         Estimate Std.Err z.value p.value
#> sigma_b   1.1076  0.0471 23.5346 <0.0001
#>
#>
#> Log-Likelihood: -2307.7 
```

`print()` leads with the association parameter `alpha` and its hazard
ratio: here, each unit increase in log bilirubin multiplies the hazard of
death by about 3.6. `summary(fit_w)` gives the full tables.

By default this fits a single random intercept (`random_effects =
"intercept"`). Maximum likelihood also supports a correlated random
intercept and slope (`random_effects = "intercept_slope"`), with a 2D
adaptive Gauss-Hermite implementation cross-validated against R's `JM`
package to within a fraction of a standard error - see
`vignette("jmjax-validation")`.

## Fitting: flexible baseline hazard, correlated random slope

For a data-driven baseline hazard shape and/or a correlated random
intercept and slope, use `jm_bayes()`. Its default spline baseline gives
the most flexibility:

```{r, eval = FALSE}
fit_s <- jm_bayes(log_bili ~ year, Surv(years, status2) ~ drug,
                  data_long = pbc2, data_surv = pbc2.id,
                  id_var = "id", time_var = "year",
                  random_effects = "intercept_slope", random_formula = ~ year,
                  chains = 2, warmup = 500, samples = 1000, seed = 1)
fit_s
#> Call:
#> jm_bayes(long_formula = log_bili ~ year, surv_formula = Surv(years, 
#>     status2) ~ drug, data_long = pbc2, data_surv = pbc2.id, id_var = "id", 
#>     time_var = "year", random_effects = "intercept_slope", random_formula = ~year, 
#>     chains = 2, warmup = 500, samples = 1000, seed = 1)
#>
#> Method: spline-PH-mcmc 
#> Precision: float64
#> Subjects: 312  | Longitudinal obs: 1945  | Events: 140 (44.9%) 
#>
#> Association (alpha): effect of the longitudinal marker on the hazard
#>   alpha = 1.1756, 95% credible interval = [1.0145, 1.3482]
#>   hazard ratio per unit of marker = 3.2401, 95% credible interval = [2.7579, 3.8505]
#>
#> Estimates by submodel:
#> Longitudinal:
#>         Estimate Std.Err CrI.lower CrI.upper   Rhat      ESS
#> beta_0    0.4911  0.0594    0.3722    0.6063 0.9994 3525.527
#> beta_1    0.1844  0.0137    0.1589    0.2112 0.9990 2691.163
#> sigma_e   0.3473  0.0067    0.3345    0.3607 1.0035 1262.061
#>
#> Survival (baseline hazard): 9 spline coefficients (W0-W8); see summary(fit) for the full table
#>
#> Survival (baseline covariates):
#>         Estimate Std.Err CrI.lower CrI.upper   Rhat      ESS
#> gamma_0  -0.0037  0.1706   -0.3433    0.3107 0.9991 4300.545
#>
#> Variance components:
#>          Estimate Std.Err CrI.lower CrI.upper   Rhat      ESS
#> sigma_b0   1.0084  0.0427    0.9291    1.0989 1.0021 465.9280
#> sigma_b1   0.1818  0.0122    0.1589    0.2081 1.0019 827.9985
#> rho        0.4143  0.0734    0.2617    0.5439 1.0027 836.6772
#>
#>
#> Convergence: max R-hat = 1.0035, min ESS = 466, divergences = 0 
#> Sampler: 2 chains x 1000 samples (500 warmup), 9.4s 
#> Rotation (control$rotate_absorbable): applied (1 direction)
```

**Always check the printed diagnostics line before trusting point
estimates.** `max R-hat` should be close to 1 (conventionally < 1.05) and
`min ESS` should be at least a few hundred for a 1000-sample fit. If
either looks off, increase `warmup`/`samples`, or - especially
for `random_effects = "intercept_slope"` at small-to-moderate sample
sizes - run more chains (`chains = 4`) and re-check,
since a single chain's own diagnostics cannot detect if it has settled
into the wrong region of a multimodal posterior (see
`vignette("jmjax-validation")` for a worked example of exactly this
check).

**For real analyses, a set of validated performance-tuning options is
recommended** on top of the basic call above - see `?jm_fit`'s
"Performance-tuning options" section, or `vignette("jmjax-validation")`'s
"Performance tuning" section for the full detail and benchmarks:

```{r, eval = FALSE}
jmjax_setup(num_devices = 4)  # once per session, enables genuinely
                               # parallel (not sequential) chains

fit_s <- jm_bayes(log_bili ~ year, Surv(years, status2) ~ drug,
                  data_long = pbc2, data_surv = pbc2.id,
                  id_var = "id", time_var = "year",
                  random_effects = "intercept_slope", random_formula = ~ year,
                  chains = 4, warmup = 500, samples = 1000,
                  control = list(spline_prior = "penalized"))
```

**The performance options that used to be listed here are now defaults or
have been withdrawn.** Earlier versions of this vignette recommended
setting `random_effects_method = "wishart_gibbs"`,
`rw2_implementation = "vectorized"`, `dense_mass_spline = TRUE` and
`progress_bar = FALSE`. A replication study (3 seeds x 2 sample sizes,
each candidate compared against its own default, with criteria written
before the results were seen) changed that picture:

| Option | Then | Now |
|---|---|---|
| `dense_mass_spline` | recommended | **default `TRUE`** - passed replication at ~2.1x faster, ~2.3x ESS/second, consistent on every run |
| `rw2_implementation` | recommended `"vectorized"` | **default `"vectorized"`** - not for speed (the original ~1.4x came from a single run and did not replicate: 0.92x ESS/second at n=1000), but because `"scan"` cannot run with the pinned jax/numpyro versions (jax-ml/jax#22045) |
| `random_effects_method` | recommended `"wishart_gibbs"` | **stays `"nuts"`** - mixed direction across seeds, and worse truth recovery, consistent with a documented inverse-Wishart small-variance bias |
| `progress_bar` | recommended `FALSE` | **default `FALSE`**, for the volume of console output rather than speed - the reported ~30x overhead did not replicate (1.00x-1.16x), so `progress_bar = TRUE` is cheap when you want it; `verbose = TRUE` gives a one-line notice instead |
| covariate standardization | manual, user-side | **automatic** - see below |

**Covariate standardization is now automatic.** jmjax centres and scales
the baseline covariates in `long_formula` internally and back-transforms
`beta` so coefficients are reported on the original scale. Nothing you
read changes; only sampling efficiency does. With an uncentred covariate
the posterior correlation between the intercept and that covariate's
coefficient is approximately `-xbar / sqrt(xbar^2 + s^2)` - measured at
`-0.976` for `age ~ N(50, 10)` - which is expensive for NUTS to traverse.
Standardizing removed it, cutting mean leapfrog steps from 228 to 50 and
raising sampling efficiency ~2.3x. Set
`control$standardize_covariates = FALSE` to disable.

**A related practical note on `time_var`'s numerical scale**: when using
`random_effects = "intercept_slope"` or `random_formula` together with a
nonlinear `long_formula` (e.g. `y ~ time + I(time^2)`), the current
adaptive-GH and NUTS backends can be sensitive to the raw numerical range
of `time_var`. If you encounter a convergence failure
(`fit$convergence$converged == FALSE`, `n_iter == 0`, or a very high
R-hat) in this configuration, try rescaling `time_var` to a moderate
range (roughly single digits - e.g. dividing "days since baseline" by
365 to get "years", or dividing "years" by 10) before refitting. This
mirrors standard practice for gradient-based optimization generally, and
is a normal part of working with adaptive-GH/HMC joint models on real
data - `JM`'s own documentation similarly notes that "difficult
datasets" for this model class may need troubleshooting (e.g. checking
sensitivity to the number of quadrature points; see
`?JM::jointModel`'s notes). See `vignette("jmjax-validation")`'s
"Time-scale sensitivity" section for a worked real-data example of
diagnosing and resolving this.

By default, `"spline-PH-mcmc"` uses:

- `spline_prior = "independent"` - independent priors on each spline
  coefficient. Set `spline_prior = "penalized"` for a P-spline (RW2 +
  Gamma precision hyperprior) that lets the data determine how smooth the
  baseline hazard should be, matching `JMbayes2`'s approach (confirmed
  hyperparameters: shape=5, rate=0.5).
- `empirical_bayes_prior = TRUE` - centers the priors for the fixed
  effects and random-effect standard deviations on an internally-fit
  `nlme::lme()` model, matching `JMbayes2`'s confirmed default behavior.
  This stabilizes estimation of the parameters most entangled with
  `alpha` (especially the random-slope variance) at small-to-moderate
  sample sizes. Set `FALSE` to use flat/uninformative priors instead.

## Extracting per-subject random effects

Only Bayesian fits (`jm_bayes()`) estimate individual subjects' random
effects directly; `jm_mle()` integrates them out with quadrature:

```{r, eval = FALSE}
head(ranef(fit_s), 3)
#>   subject_index    b0_mean     b0_sd     b1_mean      b1_sd
#> 1             1  2.1849872 0.2492208  0.21313804 0.16088919
#> 2             2 -0.4583328 0.1798143 -0.01216079 0.03420319
#> 3             3 -0.2633982 0.2034698  0.01865051 0.12799940
```

Calling `ranef()` on an MLE fit returns `NULL` with a warning, since
those methods never estimate individual `b_i` directly.

## Coefficients, intervals and predictions

A fit works with R's usual model functions. The standard errors from
`vcov()` are the ones `summary()` reports, and for MCMC fits `vcov()` and
`confint()` are computed from the posterior draws:

```{r, eval = FALSE}
confint(fit_s, "alpha")        # a 95% credible interval here
#>          2.5 %   97.5 %
#> alpha 1.014478 1.348195
round(sqrt(diag(vcov(fit_s))), 4)   # the Std.Err column of summary(fit_s)
#>   beta_0   beta_1  sigma_e    alpha       W0       W1       W2       W3 
#>   0.0594   0.0137   0.0067   0.0857   0.4603   0.5656   0.5862   0.4904 
#>       W4       W5       W6       W7       W8 sigma_b0 sigma_b1      rho 
#>   0.5323   0.5619   0.9342   1.5805   1.4532   0.0427   0.0122   0.0735 
#>  gamma_0 
#>   0.1706 

logLik(fit_w)                  # maximum-likelihood fits only
#> 'log Lik.' -2307.705 (df=8)
AIC(fit_w)
#> [1] 4631.41
BIC(fit_w)
#> [1] 4661.354
nobs(fit_w)                    # subjects, not measurements
#> [1] 312
```

For MCMC fits, `predict()` gives a subject's marker trajectory or their
probability of surviving to later times, given that they survived to
`t_from` (by default, the end of their follow-up), with posterior
intervals:

```{r, eval = FALSE}
# a patient still alive at the end of follow-up (4.1 years)
id_c <- pbc2.id$id[pbc2.id$status2 == 0 & pbc2.id$years < 8][1]
nd <- pbc2[pbc2$id == id_c, ]

predict(fit_s, newdata = nd, times = 0:4)      # log bilirubin
#>   id time  estimate     lower    upper
#> 1  5    0 0.8220302 0.4525578 1.215670
#> 2  5    1 1.2058314 0.9259460 1.496839
#> 3  5    2 1.5896327 1.3186519 1.853007
#> 4  5    3 1.9734339 1.6362631 2.310022
#> 5  5    4 2.3572352 1.9107142 2.823388
predict(fit_s, newdata = nd, process = "event", times = nd$years[1] + 0:3)
#>   id   t_from     time  estimate      lower     upper
#> 1  5 4.120578 4.120578 1.0000000 1.00000000 1.0000000
#> 2  5 4.120578 5.120578 0.7504811 0.54933290 0.8834500
#> 3  5 4.120578 6.120578 0.4899600 0.18193443 0.7446028
#> 4  5 4.120578 7.120578 0.2612318 0.01870454 0.5766713
```

Each call returns one row per time, with the posterior mean in
`estimate` and the 95% interval in `lower` and `upper`. This patient's
estimated probability of surviving two more years, given survival to the
end of follow-up, is about 0.49. This version predicts
for subjects in the fitted data; predictions for new subjects, whose
random effects would have to be estimated from their own history, are
planned.

By default a fit prints only a one-off notice while JAX compiles a large
model. `control = list(verbose = TRUE)` adds a line announcing each MCMC
run and the routine notes about warm starts and covariate
standardization; `progress_bar = TRUE` shows NumPyro's progress bar.

## Warm-starting a more expensive fit

Since `beta`, `sigma_e`, and `alpha` tend to land in similar places
regardless of baseline hazard flexibility, a fast Weibull fit's estimates
make good starting values for a more expensive spline fit. Pass them as
they are: spline coefficients, which a Weibull fit does not have, start
at zero.

```{r, eval = FALSE}
fit_spline_mle <- jm_mle(log_bili ~ year, Surv(years, status2) ~ drug,
                         data_long = pbc2, data_surv = pbc2.id,
                         id_var = "id", time_var = "year", baseline = "spline",
                         init_theta = fit_w$estimates)
coef(fit_spline_mle)[c("alpha", "gamma_0")]
#>      alpha    gamma_0 
#> 1.29438985 0.03051234 
```

## Choosing a method

| Situation | Recommended call |
|---|---|
| Quick first look, or need warm-start values | `jm_mle(baseline = "weibull")` |
| Flexible baseline hazard, frequentist | `jm_mle(baseline = "spline")` |
| Correlated random intercept + slope, simple baseline hazard | `jm_bayes(baseline = "weibull")` |
| Correlated random intercept + slope, flexible baseline hazard; subject-level predictions | `jm_bayes()` |
| Comparing against `JMbayes2` | `jm_bayes(control = list(spline_prior = "penalized"))` (the default `empirical_bayes_prior = TRUE` already matches) |

## Further reading

See `vignette("jmjax-validation")` for the full cross-validation
methodology against R's `JM` and `JMbayes2` packages, the bugs found and
fixed during development, benchmark results comparing accuracy and
sampling efficiency, and the "Performance tuning" section documenting
the validated options shown above (including an experimental option not
covered here, `random_effects_method = "wishart_gibbs_centered"`, and
its known limitations).
