---
title: "Introduction to jmjax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jmjax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What jmjax fits

A shared-parameter joint model links two submodels through a common
random effect:

- **Longitudinal submodel**: a linear mixed-effects model,
  `y_ij = X_ij %*% beta + Z_ij %*% b_i + e_ij`.
- **Survival submodel**: a relative-risk (proportional hazards) model,
  `h_i(t) = h0(t) * exp(alpha * m_i(t))`, where `m_i(t)` is the subject's
  fitted longitudinal trajectory at time `t`, and `alpha` (the
  **association parameter**) measures how strongly the longitudinal
  trajectory predicts the hazard.

jmjax offers two estimation approaches, each with its own function, and
two choices for the baseline hazard `h0(t)`:

| | Weibull baseline | Spline baseline |
|---|---|---|
| **Maximum likelihood** (adaptive Gauss-Hermite) | `jm_mle(baseline = "weibull")` | `jm_mle(baseline = "spline")` |
| **Bayesian** (NUTS) | `jm_bayes(baseline = "weibull")` | `jm_bayes(baseline = "spline")` |

Both functions take the same model arguments. Each documents only the
options that apply to it (optimizer and quadrature settings for
`jm_mle()`; chains, priors and sampler settings for `jm_bayes()`).
`jm_fit()`, which both call, remains available with the method names
`"weibull-PH-aGH"`, `"spline-PH-aGH"`, `"weibull-PH-mcmc"` and
`"spline-PH-mcmc"`.

## One-time setup

```{r, eval = FALSE}
library(jmjax)
jmjax_setup()  # creates a managed Python virtualenv with jax/numpyro
```

If you already have a Python environment with `jax`/`numpyro` installed
(e.g. a conda environment), point `reticulate` at it instead - `jmjax`
will use it automatically and skip creating its own:

```{r, eval = FALSE}
library(reticulate)
use_condaenv("my_existing_env", required = TRUE)
library(jmjax)  # will detect and use the active environment
```

## Example data

Your data needs two pieces: a long-format longitudinal data frame, and a
one-row-per-subject survival data frame, sharing a common `id` column.

```{r, eval = FALSE}
library(survival)

# data_long: one row per (subject, visit)
head(data_long)
#>   id time    y
#> 1  1  0.0 2.03
#> 2  1  0.5 2.31
#> 3  1  1.0 2.58
#> ...

# data_surv: one row per subject
head(data_surv)
#>   id time event
#> 1  1  3.2     1
#> 2  2  5.0     0
#> ...
```

## Fitting: the fast first look

Maximum likelihood with a Weibull baseline is the fastest option, has no priors to configure, and
is a good first look at your data or a source of warm-start values for a
more flexible fit later.

```{r, eval = FALSE}
fit_w <- jm_mle(
  long_formula = y ~ time,
  surv_formula = Surv(time, event) ~ 1,
  data_long = data_long,
  data_surv = data_surv,
  id_var = "id",
  time_var = "time",
  baseline = "weibull"
)
summary(fit_w)
#>                                Estimate Std.Err z.value p.value
#> beta_0 (Intercept)              2.0068  0.0241  83.29    0.0000
#> beta_1 (Slope)                  0.4599  0.0045 103.11    0.0000
#> sigma_e                         0.3586  0.0033 109.50    0.0000
#> sigma_b                         0.8975  0.0172  52.28    0.0000
#> log_lambda0                    -3.4847  0.1322 -26.36    0.0000
#> shape                           1.1163  0.0394  28.35    0.0000
#> alpha (Association)             0.6976  0.0437  15.98    0.0000
```

By default this fits a single random intercept (`random_effects =
"intercept"`). `"weibull-PH-aGH"` also supports a correlated random
intercept and slope via `random_effects = "intercept_slope"` (a 2D
adaptive Gauss-Hermite implementation, cross-validated against R's `JM`
package to within a fraction of a standard error - see
`vignette("jmjax-validation")`).

## Fitting: flexible baseline hazard, correlated random slope

For a data-driven baseline hazard shape and/or a correlated random
intercept + slope, use `jm_bayes()`. Its default spline baseline gives
the most flexibility:

```{r, eval = FALSE}
fit_s <- jm_bayes(
  long_formula = y ~ time,
  surv_formula = Surv(time, event) ~ 1,
  data_long = data_long,
  data_surv = data_surv,
  id_var = "id",
  time_var = "time",
  random_effects = "intercept_slope",
  chains = 2, warmup = 500, samples = 1000
)
print(fit_s)
#> Method: spline-PH-mcmc
#> Subjects: 500  | Longitudinal obs: 2483  | Events: 273 (54.6%)
#> Convergence: max R-hat = 1.0083, min ESS = 412, divergences = 0
```

**Always check the printed diagnostics line before trusting point
estimates.** `max R-hat` should be close to 1 (conventionally < 1.05) and
`min ESS` should be at least a few hundred for a 1000-sample fit. If
either looks off, increase `warmup`/`samples`, or - especially
for `random_effects = "intercept_slope"` at small-to-moderate sample
sizes - run more chains (`chains = 4`) and re-check,
since a single chain's own diagnostics cannot detect if it has settled
into the wrong region of a multimodal posterior (see
`vignette("jmjax-validation")` for a worked example of exactly this
check).

**For real analyses, a set of validated performance-tuning options is
recommended** on top of the basic call above - see `?jm_fit`'s
"Performance-tuning options" section, or `vignette("jmjax-validation")`'s
"Performance tuning" section for the full detail and benchmarks:

```{r, eval = FALSE}
jmjax_setup(num_devices = 4)  # once per session, enables genuinely
                               # parallel (not sequential) chains

fit_s <- jm_bayes(
  long_formula = y ~ time,
  surv_formula = Surv(time, event) ~ 1,
  data_long = data_long,
  data_surv = data_surv,
  id_var = "id",
  time_var = "time",
  random_effects = "intercept_slope",
  chains = 4, warmup = 500, samples = 1000,
  control = list(spline_prior = "penalized")
)
```

**The performance options that used to be listed here are now defaults or
have been withdrawn.** Earlier versions of this vignette recommended
setting `random_effects_method = "wishart_gibbs"`,
`rw2_implementation = "vectorized"`, `dense_mass_spline = TRUE` and
`progress_bar = FALSE`. A replication study (3 seeds x 2 sample sizes,
each candidate compared against its own default, with criteria written
before the results were seen) changed that picture:

| Option | Then | Now |
|---|---|---|
| `dense_mass_spline` | recommended | **default `TRUE`** - passed replication at ~2.1x faster, ~2.3x ESS/second, consistent on every run |
| `rw2_implementation` | recommended `"vectorized"` | **default `"vectorized"`** - not for speed (the original ~1.4x came from a single run and did not replicate: 0.92x ESS/second at n=1000), but because `"scan"` cannot run with the pinned jax/numpyro versions (jax-ml/jax#22045) |
| `random_effects_method` | recommended `"wishart_gibbs"` | **stays `"nuts"`** - mixed direction across seeds, and worse truth recovery, consistent with a documented inverse-Wishart small-variance bias |
| `progress_bar` | recommended `FALSE` | **default `FALSE`**, for the volume of console output rather than speed - the reported ~30x overhead did not replicate (1.00x-1.16x), so `progress_bar = TRUE` is cheap when you want it; `verbose = TRUE` gives a one-line notice instead |
| covariate standardization | manual, user-side | **automatic** - see below |

**Covariate standardization is now automatic.** jmjax centres and scales
the baseline covariates in `long_formula` internally and back-transforms
`beta` so coefficients are reported on the original scale. Nothing you
read changes; only sampling efficiency does. With an uncentred covariate
the posterior correlation between the intercept and that covariate's
coefficient is approximately `-xbar / sqrt(xbar^2 + s^2)` - measured at
`-0.976` for `age ~ N(50, 10)` - which is expensive for NUTS to traverse.
Standardizing removed it, cutting mean leapfrog steps from 228 to 50 and
raising sampling efficiency ~2.3x. Set
`control$standardize_covariates = FALSE` to disable.

**A related practical note on `time_var`'s numerical scale**: when using
`random_effects = "intercept_slope"` or `random_formula` together with a
nonlinear `long_formula` (e.g. `y ~ time + I(time^2)`), the current
adaptive-GH and NUTS backends can be sensitive to the raw numerical range
of `time_var`. If you encounter a convergence failure
(`fit$convergence$converged == FALSE`, `n_iter == 0`, or a very high
R-hat) in this configuration, try rescaling `time_var` to a moderate
range (roughly single digits - e.g. dividing "days since baseline" by
365 to get "years", or dividing "years" by 10) before refitting. This
mirrors standard practice for gradient-based optimization generally, and
is a normal part of working with adaptive-GH/HMC joint models on real
data - `JM`'s own documentation similarly notes that "difficult
datasets" for this model class may need troubleshooting (e.g. checking
sensitivity to the number of quadrature points; see
`?JM::jointModel`'s notes). See `vignette("jmjax-validation")`'s
"Time-scale sensitivity" section for a worked real-data example of
diagnosing and resolving this.

By default, `"spline-PH-mcmc"` uses:

- `spline_prior = "independent"` - independent priors on each spline
  coefficient. Set `spline_prior = "penalized"` for a P-spline (RW2 +
  Gamma precision hyperprior) that lets the data determine how smooth the
  baseline hazard should be, matching `JMbayes2`'s approach (confirmed
  hyperparameters: shape=5, rate=0.5).
- `empirical_bayes_prior = TRUE` - centers the priors for the fixed
  effects and random-effect standard deviations on an internally-fit
  `nlme::lme()` model, matching `JMbayes2`'s confirmed default behavior.
  This stabilizes estimation of the parameters most entangled with
  `alpha` (especially the random-slope variance) at small-to-moderate
  sample sizes. Set `FALSE` to use flat/uninformative priors instead.

## Extracting per-subject random effects

Only Bayesian fits (`jm_bayes()`) estimate individual subjects' random
effects directly; `jm_mle()` integrates them out with quadrature:

```{r, eval = FALSE}
head(ranef(fit_s))
#>   subject_index    b0_mean     b0_sd    b1_mean    b1_sd
#> 1             1  0.4771231 0.1102341 -0.0322198 0.045211
#> 2             2 -1.1023841 0.0987621  0.0611234 0.039872
#> ...
```

Calling `ranef()` on an MLE fit returns `NULL` with a warning, since
those methods never estimate individual `b_i` directly.

## Coefficients, intervals and predictions

A fit works with R's usual model functions. The standard errors from
`vcov()` are the ones `summary()` reports, and for MCMC fits `vcov()` and
`confint()` are computed from the posterior draws:

```{r, eval = FALSE}
coef(fit_s)
confint(fit_s, "alpha")              # a 95% credible interval here
sqrt(diag(vcov(fit_s)))              # matches summary(fit_s)$Std.Err

logLik(fit_w); AIC(fit_w); BIC(fit_w)  # maximum-likelihood fits only
nobs(fit_w)                            # the number of subjects
```

For MCMC fits, `predict()` gives a subject's marker trajectory or their
probability of surviving to later times, given that they survived to
`t_from` (by default, the end of their follow-up), with posterior
intervals:

```{r, eval = FALSE}
nd <- data_long[data_long$id == 2, ]     # a subject who was censored
predict(fit_s, newdata = nd, times = seq(0, 5, by = 0.5))
predict(fit_s, newdata = nd, process = "event")
```

Each call returns one row per time, with the posterior mean in
`estimate` and the interval in `lower` and `upper`. This version predicts
for subjects in the fitted data; predictions for new subjects, whose
random effects would have to be estimated from their own history, are
planned.

By default a fit prints only a one-off notice while JAX compiles a large
model. `control = list(verbose = TRUE)` adds a line announcing each MCMC
run and the routine notes about warm starts and covariate
standardization; `progress_bar = TRUE` shows NumPyro's progress bar.

## Warm-starting a more expensive fit

Since `beta`, `sigma_e`, and `alpha` tend to land in similar places
regardless of baseline hazard flexibility, a fast Weibull fit's estimates
make good starting values for a more expensive spline fit:

```{r, eval = FALSE}
fit_spline_mle <- jm_mle(
  long_formula = y ~ time,
  surv_formula = Surv(time, event) ~ 1,
  data_long = data_long,
  data_surv = data_surv,
  id_var = "id",
  time_var = "time",
  baseline = "spline",
  init_theta = c(
    fit_w$estimates[["beta_0"]], fit_w$estimates[["beta_1"]],
    log(fit_w$estimates[["sigma_e"]]), log(fit_w$estimates[["sigma_b"]]),
    rep(0, 9),  # zeros for the spline coefficients (no Weibull analog)
    fit_w$estimates[["alpha"]]
  )
)
```

## Choosing a method

| Situation | Recommended call |
|---|---|
| Quick first look, or need warm-start values | `jm_mle(baseline = "weibull")` |
| Flexible baseline hazard, frequentist | `jm_mle(baseline = "spline")` |
| Correlated random intercept + slope, simple baseline hazard | `jm_bayes(baseline = "weibull")` |
| Correlated random intercept + slope, flexible baseline hazard; subject-level predictions | `jm_bayes()` |
| Comparing against `JMbayes2` | `jm_bayes(control = list(spline_prior = "penalized"))` (the default `empirical_bayes_prior = TRUE` already matches) |

## Further reading

See `vignette("jmjax-validation")` for the full cross-validation
methodology against R's `JM` and `JMbayes2` packages, the bugs found and
fixed during development, benchmark results comparing accuracy and
sampling efficiency, and the "Performance tuning" section documenting
the validated options shown above (including an experimental option not
covered here, `random_effects_method = "wishart_gibbs_centered"`, and
its known limitations).
